Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2022-02-25 write-up/letters/response-ref-1-v1.tex

I am incredibly grateful for this review. (PRAISE OUT THE WAZZOOO).

\begin{enumerate}
\item {\bf Section 1: Introduction.} {\it It may be useful in the introduction to briefly discuss the various purposes of testing for diseases – for COVID we have largely focused on diagnostic testing of people who are symptomatic or have a known or suspect exposure. Diagnostic testing will never provide valid prevalence or incidence estimates, especially for a disease where a significant proportion of cases are asymptomatic or pauci-symptomatic and thus unlikely to present for diagnostic testing. Screening tests are more appropriate as a potential tool for reconstructing prevalence or incidence tests, although these are seldom conducted at random. The random and non-random testing samples appear to represent screening and diagnostic testing goals, respectively, and this is part of why they can be expected not to deliver the same prevalence estimates.}

\vspace{5mm}
In the introduction, we now connect random testing to its purpose as a screening tool and non-random testing which generates the case count data as a diagnostic tool.  Given the infrequent nature of random testing and desire for ongoing prevalence estimates, we motivate the second objective of our paper by the question of whether randomized screening tests can be used to address selection bias in the non-random diagnostic testing to estimate prevalence or incidence over time.
\vspace{5mm}

\item {\bf Section 1.1}
\begin{itemize}
	\item {\it For the first goal of the paper, it might be helpful to link the biases with those outlined in a set of papers from the Harvard Center for Disease Dynamics: \url{https://doi.org/10.1007/s10654-021-00727-7} and DOI:\url{10.1093/aje/kwaa188}}
	\vspace{5mm}

	We now connect with this literature on potential sources of biases in Section 1.1.  We highlight that our work adds a new perspective on this critical issue, from the error decomposition perspective of Meng et al. (2018).
	\vspace{5mm}

	\item {\it For the second goal of the paper, it would be useful to discuss how the proposed method differs from the more traditional epidemiologic method of obtaining a validation sample. As I see it, the chief distinction is that here the “gold standard” is not a different test (although one could argue that the gold standard being used here is a screening test), but rather it is a less biased selection mechanism. Clarifying the links with validation studies could be helpful for epidemiologists \& statisticians alike, as the use of a validation study to correct for selection bias is a fairly novel suggestion.}
	\vspace{5mm}

	Remark within validation subsection to contrast with validation sample. We state that this is a fairly novel
	\vspace{5mm}
	\item {\it Finally, consider linking the arguments you are making with recent work on ‘target validity’ (see: \url{https://doi.org/10.1093/aje/kwy228}). Literature on target validity has largely been focused on specifying the problem of tackling internal and external validity together, but the method you propose here is a concrete attempt to address both types of validity and extends this conversation nicely.}
	\vspace{5mm}
	Validation subsection
	\vspace{5mm}
\end{itemize}
\item {\bf Section 2.1: Diagnostic testing}
\begin{itemize}
	\item {\it Please clarify in the first paragraph the distinction between incubation period (time to symptom onset) and latent period (time to infectiousness). The viral load will be detectable by at least the end of the latent period which for SARS-CoV-2 occurs before the end of the incubation period. However, individuals are unlikely to present for diagnostic testing until after the end of the incubation period.}
	\vspace{5mm}

### 2022-02-25 write-up/covid-writeup.tex
Section 1.1 cleaning up

 This article discusses the relationship between three statistical concepts: selection bias, measurement error, and population size.  While the impact of measurement error~\citep{Smeden2019} and selection bias~\citep{Keiding2016} on estimation are both well-studied topics, here we provide a new perspective by building on the work of \cite{Meng2018} who studied an error decomposition to understand the relationship between selection bias and population size.  Specifically, we quantify the interaction between measurement-error and selection bias on statistical error, showing how the sign and magnitude can change drastically.
 % the impact on the effective sample size can therefore be quite large.

>>>>

 This article discusses the relationship between three statistical concepts: selection bias, measurement error, and population size.  Potential biases in observational studies of COVID-19 have been identified elsewhere in the literature~\cite{Kahn2021,Accorsi2020}.  While the impact of measurement error~\citep{Smeden2019} and selection bias~\citep{Keiding2016} on estimation are both well-studied topics in general, here we provide a new perspective by building on the work of \cite{Meng2018} who studied an error decomposition to understand the relationship between selection bias and population size.  Specifically, we quantify the interaction between measurement-error and selection bias on statistical error, showing how the sign and magnitude can change drastically.  We then discuss this relationship in the context of observational COVID-19 case-count data, showing the impact on the effective sample size can be quite large.
